---
title: "UAS monitoring of Cheesman Reservoir tree planting growth and survival"
subtitle: ""
author: "George Woolsey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: hide
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding){ 
    out_dir <- '../';
    rmarkdown::render(inputFile, encoding = encoding, output_file=file.path(dirname(inputFile), out_dir, 'index.html'));
    file.copy(from = "..//index.html", to = '..///data//uas_lidar_n1.html', overwrite = TRUE)
  })
---

The 2002 Hayman Fire ([Graham 2003](https://doi.org/10.2737/RMRS-GTR-114); [Moriarty & Cheng, 2012](https://scholar.google.com/scholar?cluster=7513861891704405356&hl=en&as_sdt=0,6)) was the largest in Colorado history at the time and it severely impacted the Cheesman Reservoir and watershed. The reservoir, along with ~8,000 ac surrounding, is managed by Denver Water and is adjacent to the Pike San Isabel National Forest. The high severity fire resulted in near total mortality of the ponderosa pine and Douglas-fir forest surrounding the reservoir which necessitated active planting of seedlings through partnerships between Denver Water and the US Forest Service, Colorado State Forest Service, National Resource Conservation Service and the Colorado Forest Restoration Institute through the *From Forests to Faucets* program ([Denver Water](https://storymaps.arcgis.com/stories/6ef2af96207046baa8451cf20def46cb)). 

The objective of this project is to demonstrate the use of UAS remote sensing to evaluate the survival and growth of the reforestation efforts surrounding the Cheesman Reservoir. An exploratory site visit was made on April 03, 2026 by representatives from Denver Water (Jessica Jackman), CSFS (Ethan Bucholz, Andy Whelan, and Spencer Weston), CFRI (Caden Chamberlain), and CSU (George Woolsey). A single, preliminary UAS flight was completed with the [DJI Mavic 3M](https://ag.dji.com/mavic-3-m/specs) over one of the sites that had some of the first plantings. The flight took ~10 minutes with flight settings programmed to capture data for SfM processing to generate aerial point cloud data and a spectral orthomosaic.

```{r, include=FALSE, warning=F, message=F}
################################################
##```{r,eval=T
################################################
library(knitr)
library(mapview)
# clean session
remove(list = ls())
gc()
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results = 'hide'
  , fig.width = 10.5
  , fig.height = 7
)
# option to put satellite imagery as base layer of mapview maps
  mapview::mapviewOptions(
    homebutton = FALSE
    # , basemaps = c("Esri.WorldImagery","OpenStreetMap")
    , basemaps = c("OpenStreetMap", "Esri.WorldImagery")
  )
```

# Preliminary Data

let's check out the preliminary data captured from a single UAS flight that was completed on April 03, 2026 with the DJI Mavic xx over one of the sites that had some of the first plantings. The flight took ~10 minutes with flight settings programmed to capture data for SfM processing to generate aerial point cloud data and a spectral orthomosaic.

load the standard libraries we use to do work 

```{r}
# bread-and-butter
library(tidyverse) # the tidyverse
library(viridis) # viridis colors
library(harrypotter) # hp colors
library(RColorBrewer) # brewer colors
library(scales) # work with number and plot scales
library(latex2exp)

# visualization
library(mapview) # interactive html maps
library(kableExtra) # tables
library(patchwork) # combine plots
library(ggnewscale) # new scale
library(ggrepel) # repel labels
library(GGally) # plot matrix

# spatial analysis
library(terra) # raster
library(sf) # simple features
library(lidR) # lidR
library(cloud2trees) # cloud2trees
library(exactextractr) # raster value extract
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

Though not necessary for `cloud2trees` data processing, let's quickly check out the location and structure of the data we have

```{r,eval=T,purl=FALSE}
# directory with the downloaded .las|.laz files
point_cld_folder <- "../data/CheesmanLAS/"
# is there data?
list.files(point_cld_folder, pattern = ".*\\.(laz|las)$") %>% length()
# what files are in here?
list.files(point_cld_folder, pattern = ".*\\.(laz|las)$")[1]
```

again, this is not necessary for `cloud2trees` data processing but we can use `lidR` to read the point cloud folder as a catalog which doesn't read in the actual points but just the point cloud header data which includes information on things like the spatial location of the data, the point density, and other point attributes

```{r,eval=T,purl=FALSE}
# read folder as LAScatalog
ctg_temp <- lidR::readLAScatalog(point_cld_folder)
# what information do we get about the point cloud?
ctg_temp
```

let's look at the point cloud extent on a map to orient ourselves in space

```{r,eval=T,purl=FALSE}
ctg_temp %>% 
  cloud2trees:::check_las_ctg_empty() %>% 
  purrr::pluck("data") %>% 
  mapview::mapview(popup = F, layer.name = "point cloud tile")
```

i told you that we didn't need to do any of that for `cloud2trees` data processing and to prove it, we'll remove the `ctg_temp` object from our session

```{r, warning=FALSE, message=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

we also got an orthomosaic of at least RGB spectral bands. we might also have gotten 4-band imagery if the Mavic xx used for the flight included a sensor that also captures the NIR band?

```{r}
#### pc extent vs rgb rast
rgb_rast_fnm <- "../data/CheesmanOrtho/result.tif"
rgb_rast <- terra::rast(rgb_rast_fnm)
```

what is this raster data?

```{r}
rgb_rast
```

it looks like there might actually be a fourth band...

```{r}
terra::hist(rgb_rast)
```

...or not. that fourth band is probably the alpha channel. let's check out the RGB

```{r}
rgb_rast <- rgb_rast %>% terra::subset(c(1:3))
# get them from the cloud2trees hidden package
# plot
terra::plotRGB(rgb_rast)
# terra::plot(
#   rgb_rast
#   , nc = 3
#   # , nr = 3
#   , mar = c(0.5,0.5,0.5,0.5)
#   , axes = FALSE
#   , legend = F
#   # , col = grDevices::gray.colors(n=111)
# )
```

looks pretty good and there are certainly shadows on the ground from trees (hopefully). let's zoom in on the 5000 square meter central area

```{r, fig.height=8, fig.width=8}
rgb_rast %>%
  terra::crop(
    terra::ext(rgb_rast[[1]]) %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      sf::st_centroid() %>% 
      sf::st_buffer(sqrt(5000)/2) %>% 
      sf::st_bbox() %>% 
      sf::st_as_sfc() %>% 
      sf::st_set_crs(terra::crs(rgb_rast)) %>% 
      terra::vect()
    , mask = T
  ) %>% 
  terra::plotRGB()
```

green trees!

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Point Cloud to Raster: `cloud2raster()`

Although the `cloud2trees::cloud2trees()` function combines methods in the `cloud2trees` package for an all-in-one approach, we'll instead use the `cloud2trees::cloud2raster()` function to generate a CHM from the point cloud that we can work with to perform both individual tree detection and slash pile identification (discussed later).

We'll set the options in the function to generate a CHM which represents a DSM with the ground removed and no other filtering. This high resolution (i.e. fine-grain) CHM will serve as the foundation for tree detection and slash pile detection as we can manipulate it to optimize the processing for both methods.

```{r}
# outdir
c2t_output_dir <- "../data"
c2t_process_dir <- file.path(c2t_output_dir, "point_cloud_processing_delivery")
c2t_tracking_fnm <- file.path(c2t_process_dir, "processed_tracking_data.csv")
##############################################################
# cloud2trees::cloud2raster
##############################################################
if(
  !file.exists( file.path(c2t_process_dir, "chm_0.1m.tif") )
  || !file.exists( file.path(c2t_process_dir, "dtm_0.25m.tif") )
){
  # time it
  st_temp <- Sys.time()
  # run it
  # cloud2trees
  cloud2raster_ans <- cloud2trees::cloud2raster(
    output_dir = c2t_output_dir
    , input_las_dir = point_cld_folder
    , accuracy_level = 2
    , keep_intrmdt = T
    , dtm_res_m = 0.25
    , chm_res_m = 0.1
    , min_height = 0 # effectively generates a DSM based on non-ground points
  )
  
  # timer
  mins_temp <- difftime(Sys.time(),st_temp,units = "mins") %>% as.numeric()
  # save tracking
  dplyr::tibble(
    timer_cloud2raster_mins = mins_temp
  ) %>% 
    write.csv(
      file = c2t_tracking_fnm
      , row.names = F, append = F
    )
}else{
  dtm_temp <- terra::rast( file.path(c2t_process_dir, "dtm_0.25m.tif") )
  chm_temp <- terra::rast( file.path(c2t_process_dir, "chm_0.1m.tif") )
  
  cloud2raster_ans <- list(
    "dtm_rast" = dtm_temp
    , "chm_rast" = chm_temp
  )
}
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
remove(f)
gc()
```

let's see what we got from `cloud2trees::cloud2raster()`

```{r}
cloud2raster_ans %>% names()
```

there's a DTM

```{r}
# plot to check out the fine-resolution DTM raster
cloud2raster_ans$dtm_rast %>% 
  terra::plot(col = harrypotter::hp(n=100, option = "mischief"), main = "DTM (m)", axes = F)
```

we'll use the DTM + RGB extent to create a stand polygon since we don't have a stand poly and want to use it to filter data for edge effects (i.e. remove data along edges of data collection extent)

```{r, results=F}
stand_sf <- 
  cloud2raster_ans$dtm_rast %>% 
  terra::not.na(falseNA=T) %>% 
  terra::as.polygons(values = TRUE, aggregate = TRUE) %>% 
  sf::st_as_sf() %>% 
  sf::st_buffer(-20)
stand_sf <- 
  stand_sf %>% 
  st_union() %>% 
  sf::st_intersection(
    rgb_rast %>% 
      sum(na.rm = T) %>% 
      terra::not.na(falseNA=T) %>% 
      terra::as.polygons(values = TRUE, aggregate = TRUE) %>% 
      sf::st_as_sf() %>% 
      sf::st_buffer(-20) %>% 
      st_union()
  ) %>% 
  sf::st_concave_hull(ratio = 1) %>% 
  sf::st_as_sf()
```

and we'll make sure the stand that we defined above includes the mass of both the data collection (RGB and DTM) while removing edges

```{r, results=F, fig.show='asis'}
cloud2raster_ans$dtm_rast %>% 
  terra::plot(col = harrypotter::hp(n=100, option = "mischief"), main = "DTM (m) + 'stand' boundary", axes = F)
stand_sf %>% 
  terra::vect() %>% 
  terra::plot(border = "blue", col = NA, lwd = 2, add = T)
```

there's also a CHM

```{r}
# plot to check out the fine-resolution CHM raster
cloud2raster_ans$chm_rast %>% 
  terra::plot(col = viridis::plasma(n=100), main = "CHM (m)", axes = F)
```

let's see some details about the CHM

```{r}
# what chm?
cloud2raster_ans$chm_rast
# what data?
cloud2raster_ans$chm_rast %>% terra::summary()
```

for tree detection, we'll aggregate the CHM to a lower resolution (i.e. coarser) to smooth out some of the fine detail that can increase the noise in the tree detection processing. This aggregation process also will speed up the processing and reduce the chances of computational memory issues.

```{r}
##################################################################
# aggregate to make raster more coarse for tree detection
##################################################################
# first, we'll borrow from the `cloud2trees` codebase to get a function to change the resolution of a raster exactly
###___________________________________________###
# adjust the resolution of a raster to be in exactly the target resolution
###___________________________________________###
if(
  !file.exists( file.path(c2t_process_dir, "chm_0.20m.tif") )
){
  agg_chm_rast <- cloud2trees:::adjust_raster_resolution(
    cloud2raster_ans$chm_rast
    , target_resolution = 0.20
    , fun = max
    , resample_method = "max"
    , ofile = file.path(c2t_process_dir, "chm_0.20m.tif")
  )
}else{
  agg_chm_rast <- terra::rast( file.path(c2t_process_dir, "chm_0.20m.tif") )
}
# what chm?
agg_chm_rast
# what data?
agg_chm_rast %>% terra::summary()
```

let's go back to the central 5000 square meter area and look at the CHM over the RGB

first, we'll plot the 0.1 m resolution CHM overlaid

```{r, fig.height=8, fig.width=8}
# aoi_temp
aoi_temp <- terra::ext(rgb_rast[[1]]) %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc() %>% 
  sf::st_centroid() %>% 
  sf::st_buffer(sqrt(5000)/2) %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc() %>% 
  sf::st_set_crs(terra::crs(rgb_rast)) %>% 
  terra::vect()
rgb_rast %>%
  terra::crop(aoi_temp, mask = T) %>% 
  terra::plotRGB(
    mar = c(0.1,0.1,1.5,5)
    , main = "0.1 m resolution CHM"
  )
cloud2raster_ans$chm_rast %>%
  # agg_chm_rast %>% 
  terra::crop(aoi_temp, mask = T) %>% 
  terra::plot(
    add = T
    , axes = F
    , col = viridis::plasma(n=100) # harrypotter::hp(n=100, option = "gryffindor")
    , alpha = 0.4
    , plg = list(title = "CHM (m)", title.cex = 0.9, shrink = 0.6)
  )
```

now, we'll plot the 0.2 m resolution CHM overlaid

```{r, fig.height=8, fig.width=8}
rgb_rast %>%
  terra::crop(aoi_temp, mask = T) %>% 
  terra::plotRGB(
    mar = c(0.1,0.1,1.5,5)
    , main = "0.2 m resolution CHM"
  )
agg_chm_rast %>% 
  terra::crop(aoi_temp, mask = T) %>% 
  terra::plot(
    add = T
    , axes = F
    , col = viridis::plasma(n=100) # harrypotter::hp(n=100, option = "gryffindor")
    , alpha = 0.4
    , plg = list(title = "CHM (m)", title.cex = 0.9, shrink = 0.6)
  )
```

...interesting

a few initial observations and thoughts:

* The timing of the data collection (early April) allows the green conifers to stand out against the background terrain and any dormant or senescent grasses and forbs
  - Tree segmentation could start with identifying green objects from the RGB and then confirm with structural data (or vice-versa)
* There are standing dead trees which are represented in the upper limits of the CHM (>8 m)
  - These will need to be removed in an analysis of the tree plantings via spectral filtering and/or height filtering
  - Could also filter based on some height/area ratio
* The taller tree plantings (about 1.5-4.5 meters tall) are clearly visible in the CHM
  - What window function to use for ITD from CHM to separate trees with overlapping crowns?
* Some of the shorter tree plantings (< 1.5 m) are not represented in the CHM (no CHM pixels overlapping green areas)
  - Since they are not in the CHM, it is not likely they are in the point cloud since this CHM is effectively a height-normalized DSM based on non-ground points (`cloud2raster(..., min_height = 0)`)
  - Explore identifying these potential trees first using RGB, then check the raw point cloud in these areas for segmentation and height
  - If these short trees are not represented in the point cloud, try different SfM processing settings?
  - If these short trees *are* represented in the point cloud, would need to integrate a tree segmentation method that uses point cloud data directly
  - If these short trees are not represented in the point cloud even after using different SfM processing, use the RGB for segmentation and shadows for height?

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

# Spectral-Only Segmentation

let's use the RGB orthomosaic data to attempt an Object-Based Image Analysis (OBIA) workflow to segment individual trees from the RGB imagery without machine learning. We'll do this by combining spectral indices (e.g. GRVI), local maxima filtering (really green areas of the image), and region-growing algorithms (watershed segmentation based on greenness)

some relevant research includes: 

* [Pouliot et al. (2002)](https://doi.org/10.1016/S0034-4257(02)00050-0)
* [Wang et al. (2004)](https://doi.org/10.14358/PERS.70.3.351)
* [Xu et al. (2021)](https://doi.org/10.1016/j.rse.2021.112397)


first, we'll reduce the resolution (make the raster more coarse) of the RGB raster to both smooth out the raster and reduce noise and to reduce processing time by reducing the size of the data we need to process

we'll shoot for a 0.08 m (8 cm) raster resolution compared to the `r scales::comma(terra::res(rgb_rast)[1],accuracy = 0.001)` m native resolution

```{r}
fnm_agg_rgb_rast_temp <- "../data/agg_rgb_rast.tif"
if(!file.exists(fnm_agg_rgb_rast_temp)){
  agg_rgb_rast <- cloud2trees:::adjust_raster_resolution(
    rgb_rast
    , target_resolution = 0.08
    , fun = mean
    , resample_method = "bilinear"
    , ofile = fnm_agg_rgb_rast_temp
  )  
}else{
  agg_rgb_rast <- terra::rast(fnm_agg_rgb_rast_temp)
}
# did we do it?
terra::res(agg_rgb_rast)
```

how does it look when we zoom in on a 0.5 ha area

```{r}
aoi <- terra::ext(rgb_rast[[1]]) %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc() %>% 
  sf::st_centroid() %>% 
  sf::st_buffer(sqrt(5000)/2) %>% 
  sf::st_bbox() %>% 
  sf::st_as_sfc() %>% 
  sf::st_set_crs(terra::crs(rgb_rast)) %>% 
  terra::vect()
agg_rgb_rast %>% 
  terra::crop(aoi, mask = T) %>% 
  terra::plotRGB()
```

## Spectral Indices

ok

`cloud2trees` has (currently) hidden functions for working with RGB spectral data. let's get some relevant spectral indices

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

```{r}
fnm_rgb_indices_rast_temp <- "../data/rgb_indices_rast.tif"
if(!file.exists(fnm_rgb_indices_rast_temp)){
  rgb_indices_rast <- cloud2trees:::calculate_rgb_indices(rgb_rast = agg_rgb_rast, red_band_idx = 1, green_band_idx = 2, blue_band_idx = 3)
  terra::writeRaster(rgb_indices_rast, fnm_rgb_indices_rast_temp, overwrite=T)
}else{
  rgb_indices_rast <- terra::rast(fnm_rgb_indices_rast_temp)
}
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

let's look at those indices in our central 0.5 ha area

```{r}
# did we do it?
# terra::plot(rgb_indices_rast)
rgb_indices_rast %>% 
  terra::crop(aoi, mask = T) %>% 
  terra::plot(
    nc = 3
    , col = grDevices::gray.colors(111, start = 0, end = 1)
    , mar = c(0.5,0.5,2,0.5)
    , axes = FALSE
    , legend = F
  )
```

wow. a few of those indices really make the green areas stand out (which is why they are included in `cloud2trees`). let's mask for only the really green areas using the ExGR index threshold of zero (i.e. if ExGR > 0, then likely tree) see [Meyer & Neto (2008)](https://doi.org/10.1016/j.compag.2008.03.009) and [Riehle et al. (2020)](https://doi.org/10.1016/j.compag.2019.105201)

```{r}
grvi_thrshld_rast <- terra::ifel(rgb_indices_rast$exgr > 0, rgb_indices_rast$exgr, NA)
```

plot those green pixels on top of the original RGB

```{r, include = F, eval=FALSE}
graphics::par(mfrow = c(1, 2), mar = c(4, 4, 3, 2))
terra::plotRGB(agg_rgb_rast %>% terra::crop(aoi,mask=T), main = "RGB Raster")
terra::plot(
  grvi_thrshld_rast %>% terra::crop(aoi,mask=T)
  , col =  grDevices::colorRampPalette( RColorBrewer::brewer.pal(9, "Greens")[2:9] )(100)
  , main = "ExGR > 0"
  , axes = F
  , plg = list(
    title = "ExGR"
    , cex = 0.8, title.cex = 0.5, shrink = 0.6
  )
)
graphics::par(mfrow = c(1, 1))

```

```{r}
terra::plotRGB(agg_rgb_rast %>% terra::crop(aoi,mask=T))
terra::plot(
  grvi_thrshld_rast %>% terra::crop(aoi,mask=T)
  , col =  grDevices::colorRampPalette( RColorBrewer::brewer.pal(9, "Greens")[2:9] )(100)
  # , main = "ExGR > 0"
  , axes = F
  , legend = F
  , alpha = 0.4
  # , plg = list(
  #   title = "ExGR"
  #   , cex = 0.8, title.cex = 0.5, shrink = 0.6
  # )
  , add = T
)
```

## Segment Trees

looking good. let's use `lidR::locate_trees` with a LMF (local maximum filter) using a search window sized to 80% of a mature tree crown (we don't expect mature trees here) to detect the greenest pixel (as the center of the tree) to use in a MCWS ;D

```{r}
ws_temp <- 6*0.8 # 6 = mature PIPO crown diameter
# identify tree tops
tree_tops_temp <- 
  lidR::locate_trees(
    grvi_thrshld_rast
    , algorithm = lidR::lmf(
      ws = ws_temp
      , hmin = 0 # ExGR threshold
    )
  )
```

check out locations of the greenest pixel per potential tree location

```{r}
terra::plot(
  grvi_thrshld_rast %>% terra::crop(aoi,mask=T)
  , col =  grDevices::colorRampPalette( RColorBrewer::brewer.pal(9, "Greens")[2:9] )(100)
  , axes = F
  , plg = list(
    title = "ExGR"
    , cex = 0.8, title.cex = 0.5, shrink = 0.6
  )
)
tree_tops_temp %>% 
  # drops z values
  sf::st_zm(drop = T) %>% 
  terra::vect() %>% 
  terra::crop(aoi) %>% 
  terra::plot(add = T, cex = 0.5, col = "magenta")
```

delineate crowns using a marker controlled watershed algorithm (`ForestTools::mcws()`)

```{r}
min_area_m2_temp <- pi*(0.7/3.281)^2 # 0.7 ft min radius
# delineate crowns
crowns_sf_spectral <- ForestTools::mcws(
  treetops = sf::st_zm(tree_tops_temp, drop = T) # drops z values
  , CHM = grvi_thrshld_rast
  , minHeight = 0
) %>% 
  # convert raster to polygons for each individual crown
  terra::as.polygons() %>%
  # fix polygon validity
  terra::makeValid() %>%
  # reduce the number of nodes in geometries
  terra::simplifyGeom() %>%
  # remove holes in polygons
  terra::fillHoles() %>%
  # convert to sf
  sf::st_as_sf() %>%
  dplyr::rename(treeID = 1) %>%
  # get the crown area
  dplyr::mutate(
    crown_area_m2 = as.numeric(sf::st_area(.))
  ) %>%
  #remove super small crowns
  dplyr::filter(
    crown_area_m2 > min_area_m2_temp
  ) %>% 
  # keep only the largest portion/polygon of each segment if split across non-connected pieces
  cloud2trees::simplify_multipolygon_crowns()
```

where are these?

```{r}
crowns_sf_spectral %>% dplyr::glimpse()
```

let's check out the segmented crowns (blue) with the potential tree point locations (pink) on the GRVI raster. note that we only kept crowns that met a minimum area threshold (`r scales::comma(min_area_m2_temp, accuracy = 0.01)` square meters, approx. `r scales::comma(sqrt(min_area_m2_temp/pi)*3.281, accuracy = 0.01)` ft radius)

```{r}
terra::plot(
  grvi_thrshld_rast %>% terra::crop(aoi,mask=T)
  , col =  grDevices::colorRampPalette( RColorBrewer::brewer.pal(9, "Greens")[2:9] )(100)
  , axes = F
  , plg = list(
    title = "ExGR"
    , cex = 0.8, title.cex = 0.5, shrink = 0.6
  )
)
tree_tops_temp %>% 
  # drops z values
  sf::st_zm(drop = T) %>% 
  terra::vect() %>% 
  terra::crop(aoi) %>% 
  terra::plot(add = T, cex = 0.5, col = "magenta")
crowns_sf_spectral %>% 
  terra::vect() %>% 
  terra::crop(aoi) %>% 
  terra::plot(add = T, col = NA, border = "cyan", lwd = 1.3)
```

look at the crowns only on the RGB

```{r}
terra::plotRGB(agg_rgb_rast %>% terra::crop(aoi,mask=T))
# tree_tops_temp %>% 
#   # drops z values
#   sf::st_zm(drop = T) %>% 
#   terra::vect() %>% 
#   terra::crop(aoi) %>% 
#   terra::plot(add = T, cex = 0.5, col = "magenta")
crowns_sf_spectral %>% 
  terra::vect() %>% 
  terra::crop(aoi) %>% 
  terra::plot(add = T, col = NA, border = "cyan", lwd = 1.3)
```

```{r, include=F, eval=F}
crowns %>% 
  terra::vect() %>% 
  terra::crop(aoi) %>% 
  mapview::mapview(layer.name="asdf")

terra::plotRGB(rgb_rast %>% terra::crop(crowns %>% dplyr::filter(treeID==1449) %>% sf::st_buffer(2) %>% terra::vect()) )
crowns %>% 
  dplyr::filter(treeID==1449) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = NA, border = "cyan", lwd = 1.3)


###################################################################################################
################################################################################################### 
# [Pouliot et al. (2002)](https://doi.org/10.1016/S0034-4257(02)00050-0), p 328
# Without angle removal, the delineated polygon appears to be most similar to a natural crown shape, 
# but it has some extended points and rapid indentations that do not match the crown boundary. 
# Applying the 40 degree polygon angle removal reduces the presence of these points providing 
# a generalized representation of the crown shape. Determination of the most suitable angle removal 
# is left to the user; in this study angles of less than 20 degrees were removed. However, 
# if trees have large branches, it may be necessary to include polygon angles of less than 20 degrees
# in order to make rapid directional changes in following the crown boundary.
###################################################################################################
###################################################################################################

# clean a single matrix of coordinates (handles individual rings)
remove_sharp_angles_matrix <- function(coord_matrix, min_angle_deg = 20) {
  # Close check (sf polygons repeat the first point at the end)
  # Remove the duplicate end point temporarily for sequential calculations
  n <- nrow(coord_matrix) - 1
  if (n < 3) return(coord_matrix) # Cannot form a polygon with fewer than 3 points
  
  pts <- coord_matrix[1:n, , drop = FALSE]
  keep <- rep(TRUE, n)
  
  # Calculate angles for each vertex
  for (i in 1:n) {
    # Get previous, current, and next indices (with wrap-around)
    prev_idx <- if (i == 1) n else i - 1
    next_idx <- if (i == n) 1 else i + 1
    
    A <- pts[prev_idx, ]
    B <- pts[i, ]       # Current vertex
    C <- pts[next_idx, ]
    
    # Create vectors from B
    v1 <- A - B
    v2 <- C - B
    
    # Calculate angle using dot product
    dot_prod <- sum(v1 * v2)
    mag1 <- sqrt(sum(v1^2))
    mag2 <- sqrt(sum(v2^2))
    
    # Prevent division by zero and NaN in acos due to floating point errors
    if (mag1 == 0 || mag2 == 0) {
      keep[i] <- FALSE
      next
    }
    
    cos_theta <- dot_prod / (mag1 * mag2)
    cos_theta <- max(-1, min(1, cos_theta)) # Clamp to [-1, 1]
    
    angle_deg <- acos(cos_theta) * (180 / pi)
    
    # If angle is sharper than the threshold, mark for removal
    if (angle_deg < min_angle_deg) {
      keep[i] <- FALSE
    }
  }
  
  # Filter points
  cleaned_pts <- pts[keep, , drop = FALSE]
  
  # If we destroyed the polygon structure (fewer than 3 points remaining), return empty
  if (nrow(cleaned_pts) < 3) {
    return(matrix(numeric(0), ncol = 2))
  }
  
  # Re-close the polygon ring by appending the first point to the end
  cleaned_matrix <- rbind(cleaned_pts, cleaned_pts[1, ])
  return(cleaned_matrix)
}

# 2. Function to process an sfg (geometry object) supporting POLYGON and MULTIPOLYGON
clean_geometry_angles <- function(geom, min_angle_deg = 20) {
  if (st_is_empty(geom)) return(geom)
  
  geom_type <- as.character(st_geometry_type(geom))
  
  if (geom_type == "POLYGON") {
    # Loop through exterior and interior rings
    cleaned_rings <- lapply(geom, remove_sharp_angles_matrix, min_angle_deg = min_angle_deg)
    # Filter out any rings that became invalid/empty
    cleaned_rings <- cleaned_rings[sapply(cleaned_rings, nrow) >= 4]
    if (length(cleaned_rings) == 0) return(st_polygon())
    return(st_polygon(cleaned_rings))
    
  } else if (geom_type == "MULTIPOLYGON") {
    cleaned_polys <- lapply(geom, function(poly) {
      cleaned_rings <- lapply(poly, remove_sharp_angles_matrix, min_angle_deg = min_angle_deg)
      cleaned_rings[sapply(cleaned_rings, nrow) >= 4]
    })
    # Filter out empty polygons
    cleaned_polys <- cleaned_polys[sapply(cleaned_polys, length) > 0]
    if (length(cleaned_polys) == 0) return(st_multipolygon())
    return(st_multipolygon(cleaned_polys))
    
  } else {
    # Return non-polygon geometries unchanged
    return(geom)
  }
}

# 3. Main wrapper function for your sf dataframe
clean_polygon_angles <- function(sf_data, min_angle_deg = 20) {
  # Ensure input is an sf object
  if (!inherits(sf_data, "sf")) stop("Input must be an sf object")
  
  # Extract geometry column name
  geom_col <- attr(sf_data, "sf_column")
  
  # Apply cleaning function to every row's geometry
  sf_data[[geom_col]] <- st_sfc(lapply(st_geometry(sf_data), clean_geometry_angles, min_angle_deg = min_angle_deg), 
                                crs = st_crs(sf_data))
  
  # Remove rows that became empty geometries during the process
  sf_data <- sf_data[!st_is_empty(sf_data), ]
  
  # Fix any self-intersections caused by node removal using st_make_valid
  sf_data <- st_make_valid(sf_data)
  
  return(sf_data)
}

```

## Identify Tree Height

let's use the CHM data to get the tree height and tree center point for each segmented crown polygon. we could also refine the crown polygon shape by keeping only areas within the RGB-delineated crown that have a CHM height > 0 to remove portions that may have been included due to spectral noise, bleeding, or grass...

```{r}
# get the max chm value and the xy of the max value from the chm raster for each crown
chm_df_temp <- 
  exactextractr::exact_extract(
    x = cloud2raster_ans$chm_rast
    , y = crowns_sf_spectral %>% sf::st_transform(terra::crs(cloud2raster_ans$chm_rast))
    , summarize_df = T
    , include_xy = T
    , fun = #"max"
      function(dta) {
        valid_dta <- dta[!is.na(dta$value), ]
        # index of the maximum value
        max_idx <- which.max(valid_dta$value)
        # get the max value, x coordinate, and y coordinate 
        dplyr::tibble(
          max_val = valid_dta$value[max_idx]
          , x_coord = valid_dta$x[max_idx]
          , y_coord = valid_dta$y[max_idx]
        )
      }
    # , default_value = terra::minmax(cloud2raster_ans$chm_rast)[1]
    , progress = F
    , append_cols = "treeID" # names(crowns)
    , force_df = T
    # , colname_fun = function(fun_name, values, weights, ...) {
    #   paste0(fun_name, "_", values)
    # }
  ) %>% 
  dplyr::rename(
    tree_height_m = max_val
    , tree_x = x_coord
    , tree_y = y_coord
  )
# chm_df_temp %>% dplyr::glimpse()
# add to tree list and clean missing
crowns_sf_spectral <-
  crowns_sf_spectral %>%
  dplyr::ungroup() %>% 
  dplyr::left_join(chm_df_temp, by = "treeID") %>% 
  dplyr::mutate(
    # if no chm value, set ht to minimum chm value
    tree_height_m = dplyr::coalesce(
        tree_height_m
        , terra::minmax(cloud2raster_ans$chm_rast)[1]
      ) %>% 
      dplyr::if_else(.<0,0,.)
    , tree_height_ft = tree_height_m*3.281
    # just the center of the segment if no chm value
    , tree_x = dplyr::coalesce(
        tree_x
        , sf::st_coordinates(sf::st_point_on_surface(.))[,1]
      )
    , tree_y = dplyr::coalesce(
        tree_y
        , sf::st_coordinates(sf::st_point_on_surface(.))[,2]
      )
  )
# get point data to represent tree tops 
treetops_sf_spectral <-
  crowns_sf_spectral %>% 
  sf::st_drop_geometry() %>% 
  sf::st_as_sf(coords = c("tree_x", "tree_y"), crs = sf::st_crs(crowns_sf_spectral)) %>% 
  # filter for only those in the stand
  sf::st_intersection(
    stand_sf %>% dplyr::mutate(is_in_stand = T) %>% dplyr::select(is_in_stand)
  )
# treetops_sf_spectral %>% dplyr::glimpse()
# filter crowns based on the intersected points
crowns_sf_spectral <- crowns_sf_spectral %>% 
  dplyr::inner_join(
    treetops_sf_spectral %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(treeID,is_in_stand)
    , by = "treeID"
  )
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

what do we have now on our crowns data?

```{r}
crowns_sf_spectral %>% dplyr::glimpse()
```

look at the distribution of tree heights

```{r}
crowns_sf_spectral %>% 
  sf::st_drop_geometry() %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(x = tree_height_ft)) +
  ggplot2::geom_density(fill = "plum", color = "plum", alpha = 0.2) +
  ggplot2::scale_x_continuous(breaks = scales::breaks_extended(11)) +
  ggplot2::labs(x = "tree height (ft)", y = "") +
  ggplot2::theme_light() +
  ggplot2::theme(axis.text.y = ggplot2::element_blank(), axis.ticks.y = ggplot2::element_blank())
```

summary of values

```{r, echo=FALSE, message=FALSE}
crowns_sf_spectral %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(tidyselect::starts_with("tree_height_")) %>% 
  dplyr::mutate(dplyr::across(dplyr::everything(), ~round(.x,1))) %>% 
  summary() %>% 
  knitr::kable(format = "markdown")
```

say what? there is a tree `r scales::comma(max(crowns_sf_spectral$tree_height_ft), accuracy = 0.1)` ft tall? where?

```{r}
rgb_rast %>% 
  terra::crop(
    crowns_sf_spectral %>% 
      dplyr::slice_max(order_by = tree_height_m, n = 1, with_ties = F) %>% 
      sf::st_buffer(12) %>% 
      terra::vect()
  ) %>% 
terra::plotRGB()
# add chm
cloud2raster_ans$chm_rast %>% 
  terra::crop(
    crowns_sf_spectral %>% 
      dplyr::slice_max(order_by = tree_height_m, n = 1, with_ties = F) %>% 
      sf::st_buffer(12) %>% 
      terra::vect()
  ) %>% 
terra::plot(
  col = viridis::plasma(n=100)
  , alpha = 0.4
  , plg = list(title = "CHM (m)", title.cex = 0.9, shrink = 0.6)
  , add = T
)
#add crown
crowns_sf_spectral %>% 
  dplyr::slice_max(order_by = tree_height_m, n = 1, with_ties = F) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = NA, border = "cyan", lwd = 1.3)
```

interesting. let's get rid of the CHM so we can see what's going on

```{r}
rgb_rast %>% 
  terra::crop(
    crowns_sf_spectral %>% 
      dplyr::slice_max(order_by = tree_height_m, n = 1, with_ties = F) %>% 
      sf::st_buffer(12) %>% 
      terra::vect()
  ) %>% 
terra::plotRGB()
#add crown
crowns_sf_spectral %>% 
  dplyr::slice_max(order_by = tree_height_m, n = 1, with_ties = F) %>% 
  terra::vect() %>% 
  terra::plot(add = T, col = NA, border = "cyan", lwd = 1.3)
```

it looks like there might be some green coming up directly below a standing dead tree and not likely a surviving tall tree

what about the third tallest tree which as a height of `r crowns_sf_spectral %>% dplyr::arrange(desc(tree_height_m)) %>% dplyr::slice(3) %>% dplyr::pull(tree_height_ft) %>% scales::comma(accuracy=0.1)` ft?

```{r}
rgb_rast %>% 
  terra::crop(
    crowns_sf_spectral %>% 
      dplyr::arrange(desc(tree_height_m)) %>% 
      dplyr::slice(3) %>%
      sf::st_buffer(12) %>% 
      terra::vect()
  ) %>% 
terra::plotRGB()
#add crown
crowns_sf_spectral %>% 
  dplyr::arrange(desc(tree_height_m)) %>% 
  dplyr::slice(3) %>%  
  terra::vect() %>% 
  terra::plot(add = T, col = NA, border = "cyan", lwd = 1.3)
```

that looks like the same story where a tree is growing directly near a standing dead tree

with this knowledge of these occurrences, let's look at the distribution of tree heights across the data collection area

```{r}
agg_rgb_rast %>% 
  # terra::crop(aoi,mask=T) %>% 
  terra::plotRGB()

treetops_sf_spectral %>% 
  terra::vect() %>% 
  terra::plot(
    "tree_height_ft"
    , col = grDevices::colorRampPalette( RColorBrewer::brewer.pal(9, "YlOrBr")[2:9] )(100)
    , alpha = 0.7
    , add = T
    , pch = 16
    , cex = 0.8
    , plg = list(title = "tree ht. (ft)", title.cex = 0.9)
  )
```

neat but it's kind of hard to see. let's zoom in on that central 0.5 ha area

```{r}
agg_rgb_rast %>% 
  terra::crop(aoi,mask=T) %>%
  terra::plotRGB()

treetops_sf_spectral %>% 
  terra::vect() %>% 
  terra::crop(aoi) %>%
  terra::plot(
    "tree_height_ft"
    , col = grDevices::colorRampPalette( RColorBrewer::brewer.pal(9, "YlOrBr")[2:9] )(100)
    , alpha = 0.7
    , add = T
    , pch = 16
    # , cex = 2
    , plg = list(title = "tree ht. (ft)", title.cex = 0.9)
  )
```

finally, how many trees did we identify?

```{r}
area_ha_temp <- sf::st_area(stand_sf) %>% as.numeric() %>% `/`(10000)
dplyr::tibble(
  TPH = nrow(treetops_sf_spectral)/area_ha_temp
  , TPA = nrow(treetops_sf_spectral)/(area_ha_temp*2.471)
) %>% 
dplyr::mutate(dplyr::across(dplyr::where(is.numeric), ~scales::comma(.x,accuracy=0.1))) %>% 
tidyr::pivot_longer(dplyr::everything()) %>% 
kableExtra::kbl(
  caption = paste0(
    "Green Trees Detected over Demonstration Flight Area"
    , " ("
    , scales::comma(area_ha_temp,accuracy = 0.1,suffix = " ha")
    , " ; "
    , scales::comma((area_ha_temp*2.471),accuracy = 0.1,suffix = " ac")
    , ")"
    , "<br />"
    , scales::comma(nrow(treetops_sf_spectral),accuracy = 1,suffix = " trees")
  )
  , escape = F
  , col.names = c("","")
) %>% 
kableExtra::kable_styling()
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

# Field Data Collection Planning

now that we have established this proof-of-concept for segmenting individual trees from UAS-collected spectral data and estimating physical properties (such as height) using 3D digital aerial photogrammetry (DAP) point clouds, the next phase is to execute field data collection.

Remember, the primary objective of this project is to evaluate the success of the reforestation efforts by answering the question: *What proportion of the ~100 planted trees per acre are still alive today?*

Because counting every surviving tree across the expansive planting footprint is logistically and financially impossible, we are using UAS data to achieve a near-census of live trees. However, to trust those landscape-scale estimates, we must first use targeted field plots to validate the accuracy of our UAS workflow. 

Specifically, the field collection will be used to answer UAS-workflow validation questions across the following dimensions:

**Tree Detection Accuracy**

* Recall: What proportion of actual, ground-verified trees did the UAS method successfully detect?
* Precision: What proportion of UAS-predicted trees were actual, verified trees on the ground?
* F-score: What is the overall, balanced tree-level detection accuracy of the UAS method?

**Live Tree Density (Trees Per Acre - TPA)**

* TPA RMSE: How precise is the UAS method at measuring live tree density at the planting unit level?
* TPA Mean Error: What is the directional bias (over- or under-estimation) of the UAS density estimates compared to field-measured values?
* TPA MAPE: What is the average relative percentage error of the UAS density estimates?

**Tree-Level Physical Form**

* Height RMSE: How precise are the UAS photogrammetry height estimates compared to physical field measurements?
* Height Mean Error: What is the directional bias of the UAS height estimates?
* Height MAPE: What is the average relative percentage error of the UAS height estimates?

While plot-level field data provides exact measurements of planting survivability, those insights are constrained to the small fractions of the planting extent that ground crews actually measure. If our UAS validation demonstrates acceptable forest inventory accuracy against the plot data, we can confidently scale the UAS measurements across the entire planted footprint. Ultimately, we will rely on the UAS data to generate a comprehensive, planting-wide estimate of survivability, using the UAS-derived TPA RMSE to bound our conclusions with a statistical range of estimated survivability over the entire planted area.

The primary objective of this work is to answer the question *how successful were the reforestation planting efforts?* This can be quantified based on the survivability of the planted trees: *what proportion of the ~100 planted trees per acre are still alive?* The ideal data to answer this question would be to count all of the trees that are living at present day and compare that against how many trees were planted. However, quantifying planting success by comparing current live tree counts against original planting data is complicated by several historical and financial constraints:

1.  Planting Uncertainty: The exact number of trees originally planted is unknown.
2.  Spatial Uncertainty: Precise planting maps, boundaries, and spatial patterns were not well recorded.
3.  Methodological Shifts: Tree planting techniques changed over time.
4.  Overlapping Efforts: Several areas had multiple, compounding planting missions.
5.  Financial Constraints: Manually surveying the expansive planting area is costly and financially infeasible.
6.  Extrapolation Risks: Relying solely on traditional field plots requires extrapolating limited sample sizes over broad, heterogeneous landscapes.

To mitigate these constraints, our field data collection will prioritize a targeted, non-random sampling strategy focused on the following:

1) We will sample across as many different planting years as possible to effectively capture variations in historical planting methods and survivability trends dictated by planting age.
2) Sampling within target planting years will focus on areas with known tree survival. Because we do not know exactly where historical planting occurred, focusing on surviving clusters guarantees that field sampling takes place in areas that were actually planted. This avoids wasting field time on areas that were either never planted or suffered total mortality (which we wouldn't know because we don't know exactly were plants went in the ground).
3) We will sample a large enough area to ensure a representative count of survivors for each target year. Historical data indicates that all planting missions targeted a density of ~100 trees per acre. Therefore, an entire acre sample should ideally yield ~100 trees under 100% survivability conditions, while a half-acre sample should yield ~50 trees, and so on.

## Topography Data

Because the primary goal of this field work is to validate the UAS method, we must ensure that our validation metrics (F-score, RMSE, and MAPE) are testing the algorithm's performance across different tree structures, rather than being skewed by external environmental factors. Topography can alter UAS sensor data by introducing variable lighting geometries (shadowing, solar glare, etc.) and geometric distortions in the DAP reconstructions (e.g., on steep slopes). To isolate how well the UAS workflow performs across different tree sizes and ages (i.e. planting years), we will attempt to distribute field plots across matching topographic cohorts (slope and aspect profiles) across all target planting years. This controls for terrain-induced sensor changes.

We downloaded a 1m DEM from the [USGS National Map](https://apps.nationalmap.gov/downloader/) and here is an example DEM tile with it's [metadata](https://www.sciencebase.gov/catalog/item/69c20ed8b66b01d6ebf33cae)

We use the DEM to perform many raster calculations to get measures of slope, aspect, latitude and more. One of the primary metrics we are after is topographic heat load index (HLI) which was developed to combine slope, folded aspect (i.e., “southwestness”), and latitude to estimate terrain-driven solar heating at a given location ([McCune & Keon 2002: eq. 2](https://doi.org/10.1111/j.1654-1103.2002.tb02087.x)). Higher values of HLI near one are typical on steep southwesterly slopes, while low values near zero are common on steep northeasterly slopes. Moderate values are common in flatter terrain and on northwesterly and southeasterly aspects. 

let's read in the Cheesman

```{r}
# https://geo.colorado.edu/catalog/47540-5c93a0c3a1c5dc000989496d
fnm_temp <- "../data/LAKES/LAKES.shp"
cheesman_sf <- sf::st_read(
  fnm_temp
  # Cheesman Lake
  , query = 
    sprintf(
      "SELECT * FROM %s WHERE NAME IN (%s)"
      , basename(fnm_temp) %>% stringr::str_remove("^.*/") %>% stringr::str_remove("\\.[^.]+$")
      , stringr::str_c("'", c("Cheesman Lake"), "'", collapse = ", ")
    )
  , quiet = T
) %>% 
sf::st_union()
# cheesman_sf %>% dplyr::glimpse()
# mapview::mapview(cheesman_sf, zcol = "NAME")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Elevation

this data is directly from the USGS National Map

```{r, results=F, fig.show='asis', fig.height=9, fig.width=5.5}
# DEM covering entire area
# read in USGS 1m DEM from national map: https://apps.nationalmap.gov/downloader/
dem_fnm <- "../data/usgs_dem/dem_1m_rast.tif"

# dem_1m_rast <-
#   list.files(path = dirname(dem_fnm), pattern = "*\\ArapahoRooseveltPikeNF_D23.tif$", full.names = TRUE, recursive = F) %>%
#   .[1] %>%
#   terra::rast()
# terra::plot(dem_1m_rast)
# terra::plot(terra::vect(cheesman_sf), col = NA, border = "white", add = T)

if(!file.exists(dem_fnm)){
  dem_list <- 
    list.files(path = dirname(dem_fnm), pattern = "*\\ArapahoRooseveltPikeNF_D23.tif$", full.names = TRUE, recursive = F) %>% 
    unique() %>% 
    purrr::map(terra::rast)

  # terra::plot(dem_list[[1]])
  
  # combine the rasters into a single unified mosaic dataset
  dem_1m_rast <- terra::mosaic(terra::sprc(dem_list), fun = "mean", filename = dem_fnm, overwrite=F)
  
}else{
  dem_1m_rast <- terra::rast(dem_fnm)
}
# name
names(dem_1m_rast) <- "elevation"

# dem_1m_rast
# dem_1m_rast %>% terra::plot(col = harrypotter::hp(n=100, option = "mischief"), main = "DEM (m)", axes = F)
# dem_1m_rast %>% terra::crs()

cheesman_sf <- cheesman_sf %>% sf::st_transform(terra::crs(dem_1m_rast))

# plot
terra::plot(
  dem_1m_rast
  , col = harrypotter::hp(n=100, option = "mischief")
  , axes = F
  , mar=c(1,1,1.6,4)
  , plg = list(
     title = "elevation (m)"
     , x = "right"
     , shrink = 0.6
     , title.cex = 0.8
     , cex = 0.8
   )
)
terra::plot(terra::vect(cheesman_sf), col = NA, border = "black", add = T)

```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Terrain

now we need to get the terrain inputs for HLI which include: a raster of latitudes, slope, and aspect

we'll start with latitude. literally a raster of latitude values

```{r, results=F, fig.show='asis', fig.height=9, fig.width=5.5}
## now we need to get a raster of the WGS84 latitude values
# # extract cell coordinates and force project the to WGS84 for latitude
# to limit the data we'll process and since reprojecting lots of data with sf::sf_project(); terra::project(); or sf::st_transform() is slow
# we'll just work with the first column from the original raster, transform the y to get latitude, apply that latitude for each column by row based on the str of the original raster

# get the first column coordinates
# ?terra::xyFromCell
first_col_xy_temp <- 
  terra::xyFromCell(
    dem_1m_rast
    , cell = seq(from = 1, to = terra::ncell(dem_1m_rast), by = terra::ncol(dem_1m_rast))
  )
# head(first_col_xy_temp)
# nrow(first_col_xy_temp)

# project to wgs84 for latitude
# ?sf::sf_project
wgs84_proj_temp <- sf::sf_project(
  from = terra::crs(dem_1m_rast)
  , to = sf::st_crs(4326)
  , pts = first_col_xy_temp
)
# head(wgs84_proj_temp)
# nrow(wgs84_proj_temp)
# wgs84_proj_temp[,2] %>% head()


# make a matrix str to match original raster but each column in a given row will have the same latitude
lat_mat_temp <- dplyr::tibble(
  n = 1:nrow(wgs84_proj_temp)
  , lat = wgs84_proj_temp[,2]
)
# # matrix(
# #   wgs84_proj_temp[,2]
# #   , nrow = terra::nrow(dem_1m_rast)
# #   , ncol = terra::ncol(dem_1m_rast)
# #   , byrow = F
# # )
# # lat_mat_temp[1:5, 1:10]
# lat_mat_temp %>% head()

# ?terra::init
# terra::init(dem_1m_rast, "row") %>% terra::plot()
# template raster
row_rast_temp <- terra::init(dem_1m_rast, fun="row")

# ?terra::classify
# reclassify row numeric values to lat values
f1_temp <- file.path(dirname(dem_fnm), "lat_degrees_rast.tif" )
if(!file.exists(f1_temp)){
  lat_degrees_rast <- terra::classify(row_rast_temp, rcl = lat_mat_temp, filename = f1_temp, overwrite = F)
}else{
  lat_degrees_rast <- terra::rast(f1_temp)
}
# terra::plot(lat_degrees_rast)

# convert latitude degrees into radians
f2_temp <- file.path(dirname(dem_fnm), "lat_rad_rast.tif" )
if(!file.exists(f2_temp)){
  lat_rad_rast <- 
    terra::classify( 
      row_rast_temp
      , rcl = 
        lat_mat_temp %>% 
        dplyr::mutate(lat = lat * (pi / 180)) 
      , filename = f2_temp, overwrite = F
    )
  # # lat_rad_rast <- lat_degrees_rast * (pi / 180)
}else{
  lat_rad_rast <- terra::rast(f2_temp)
}
names(lat_rad_rast) <- "latitude_radians"
# lat_rad_rast %>% terra::plot()

# plot
terra::plot(
  lat_degrees_rast
  , col = viridis::mako(n=100)
  , axes = F
  , mar=c(1,1,1.6,4)
  , plg = list(
     title = "latitude (deg)"
     , x = "right"
     , shrink = 0.6
     , title.cex = 0.8
     , cex = 0.8
   )
)
terra::plot(terra::vect(cheesman_sf), col = NA, border = "black", add = T)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

now, let's look at slope and aspect which we can use to make a hillshade, so, from now on we can use the hillshade for plotting

```{r, results=F, fig.show='asis', fig.height=9, fig.width=5.5}
# get slope and aspect in radians
# ?terra::terrain
f3_temp <- file.path(dirname(dem_fnm), "terrain_rad_rast.tif" )
f31_temp <- file.path(dirname(dem_fnm), "slope_rad_rast.tif" )
f32_temp <- file.path(dirname(dem_fnm), "aspect_rad_rast.tif" )
if(
  !file.exists(f3_temp) ||
  !file.exists(f31_temp) ||
  !file.exists(f32_temp)
){
  terrain_rad_rast  <- terra::terrain(dem_1m_rast, v = c("slope","aspect"), unit = "radians", filename = f3_temp, overwrite = T)
  
  # because these values are smoothed, the terra::terrain result extent doesn't match with the input data...we need it to
  # align using the elevation template
  slope_fixed_temp <- terra::resample(
    terrain_rad_rast$slope
    , dem_1m_rast
    , method = "near"
    , filename = f31_temp
    , overwrite = T
  )
  
  # align using the elevation template
  aspect_fixed_temp <- terra::resample(
    terrain_rad_rast$aspect
    , dem_1m_rast
    , method = "near"
    , filename = f32_temp
    , overwrite = T
  )
  
  # combine and rename them to match the terra::terrain structure
  terrain_rad_rast <- c(
    slope_fixed_temp
    , aspect_fixed_temp
  )
  names(terrain_rad_rast) <- c("slope","aspect")
  
}else{
  # read them back 
  slope_fixed_temp <- terra::rast(f31_temp)
  aspect_fixed_temp <- terra::rast(f32_temp)
  
  # combine and rename them to match the terra::terrain structure
  terrain_rad_rast <- c(
    slope_fixed_temp
    , aspect_fixed_temp
  )
  names(terrain_rad_rast) <- c("slope","aspect")
}
# terra::plot(terrain_rad_rast$slope)
# terra::plot(terrain_rad_rast$aspect)
# identical(terra::ext(terrain_rad_rast$slope), terra::ext(dem_1m_rast))
# identical(terra::ext(terrain_rad_rast$aspect), terra::ext(dem_1m_rast))

# # hillshade?
# ?terra::shade
f4_temp <- file.path(dirname(dem_fnm), "shade_rast.tif" )
if(!file.exists(f4_temp)){
  shade_rast <- 
    terra::shade(
      slope = terrain_rad_rast$slope
      , aspect = terrain_rad_rast$aspect
      , angle = 45, direction = 315
      , filename = f4_temp, overwrite = F
    )
}else{
  shade_rast <- terra::rast(f4_temp)
}
```

slope (in radians)

```{r, results=F, fig.show='asis', fig.height=9, fig.width=5.5}
terra::plot(shade_rast, col = grDevices::gray(0:100/100), legend = F, axes = F, mar=c(1,1,1.6,4)) #B-L-T-R
terra::plot(
  terrain_rad_rast$slope
  , col = viridis::inferno(100)
  , add = T
  , alpha = 0.4
  , plg = list(
     title = "slope (rad)"
     , x = "right"
     , shrink = 0.6
     , title.cex = 0.8
     , cex = 0.8
   )
)
terra::plot(terra::vect(cheesman_sf), col = NA, border = "black", add = T)
```

aspect (in radians)

```{r, results=F, fig.show='asis', fig.height=9, fig.width=5.5}
terra::plot(shade_rast, col = grDevices::gray(0:100/100), legend = F, axes = F, mar=c(1,1,1.6,4))
terra::plot(
  terrain_rad_rast$aspect
  , col = viridis::turbo(100)
  , add = T
  , alpha = 0.5
  , plg = list(
     title = "aspect (rad)"
     , x = "right"
     , shrink = 0.6
     , title.cex = 0.8
     , cex = 0.8
   )
)
terra::plot(terra::vect(cheesman_sf), col = NA, border = "black", add = T)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### HLI

now let's get topographic heat load index (HLI) which was developed to combine slope, folded aspect (i.e., “southwestness”), and latitude to estimate terrain-driven solar heating at a given location ([McCune & Keon 2002: eq. 2](https://doi.org/10.1111/j.1654-1103.2002.tb02087.x)). Higher values of HLI near one are typical on steep southwesterly slopes, while low values near zero are common on steep northeasterly slopes. Moderate values are common in flatter terrain and on northwesterly and southeasterly aspects. 

| HLI Cluster Group | Index Range | Topographic Profile | Plant Growth & Microclimate Context |
| :--- | :--- | :--- | :--- |
| **Cool / Mesic** | `0.00` to `0.25` | Steep Northeast Slopes | Minimized solar exposure, high moisture retention, and buffered soil temperatures. Protects shade-tolerant vegetation and acts as a climate refugium. |
| **Cool-Moderate** | `> 0.25` to `0.50` | Mild North/East Slopes & Deep Ravines | Low-to-moderate solar loading. Evapotranspiration rates stable, productive vegetative growth and reliable seed germination. |
| **Warm-Moderate** | `> 0.50` to `0.75` | Flat Areas / Mild South & West Slopes | Baseline-to-high solar exposure. Typical open-sky growing conditions where photosynthesis is highly active but constrainted by precipitation and soil moisture. |
| **Hot / Xeric** | `> 0.75` to `1.00` | Steep Southwest Slopes | Maximum afternoon solar radiation. Elevated thermal stress and rapid soil drying curves create a harsh, drought-prone microclimate selecting for hardy, heat-tolerant plant species. |

define those groupings in a data frame

```{r}
hli_grp_df <- tibble::tibble(
  hli_ll = c(0,0.25,0.5,0.75)
  , hli_ul = c(0.25,0.5,0.75,1)
  , hli_grp = c("cool-mesic", "cool-moderate", "warm-moderate", "hot-xeric")
)
# hli_grp_df
```

```{r, results=F, fig.show='asis', fig.height=9, fig.width=5.5}
#########################################################
#########################################################
# make a function to get HLI
# calc topographic heat load index (HLI) 
# McCune & Keon (2002) Eq 2 
#########################################################
#########################################################
get_hli_fn <- function(latitude_radians, slope, aspect){ 
  get_hli_as_index <- T
  # folded aspect
  # # Aspect must first be ‘folded’ about the north-south line, rescaling 0-360 to 0-180
  #   # , such that NE = NW, E = W, etc.: 
  
  if(get_hli_as_index){
    # To approximate heat load, the equation can be shifted
    # from a maximum on south slopes to a maximum on southwest slopes and a minimum on northeast slopes.
    # This is accomplished by ‘folding’ the aspect about the NE - SW line
    # , such that NE becomes zero degrees and SW becomes 180:
    
    #   if aspect in degrees: Folded aspect = | 180 – |aspect – 225| |
    #   if aspect in radians: Folded aspect = | pi – |aspect – (5pi/4)| |
    
    # One then applies the identical
    # equation as for direct radiation. The result must, however, be considered a unitless index of heat load, since
    # we have no basis for converting the result into a cumulative measure of temperature, such as degree days
    
    folded_aspect <- abs( pi - abs(aspect - (5*pi/4)) )
  }else{
    # to get HLI in MJ cm^–2 yr^–1
    #   if aspect in degrees: Folded aspect = 180 – |Aspect – 180|
    #   if aspect in radians: Folded aspect = pi - |Aspect – pi|    
    folded_aspect <- pi - abs(aspect - pi)
  }

  # terra::plot(folded_aspect)
  
  # calc topographic heat load index (HLI) 
  # see: McCune & Keon (2002) Eq 2 
  # Eq. 2 increases the R2 to 0.98 by excluding slopes steeper than 60, an inconsequential omission for almost all data sets.
  # To apply the models in a spreadsheet, set up three columns with the input variables in radians.
  # This equation returns ln(Rad, MJ cm^–2 yr^–1). It can be returned to an arithmetic scale with the EXP(x) function.
  hli <- exp(
    # Eq 1
    # –1.467 + 
    # 1.582*COS(G3)*COS(H3) – 
    # 1.5*COS(I3)*SIN(H3)*SIN(G3) - 
    # 0.262*SIN(G3)*SIN(H3) + 
    # 0.607*SIN(I3)*SIN(H3)
    -1.236 + 
    ( 1.350*cos(latitude_radians)*cos(slope) ) +
    ( (-1*1.375)*cos(folded_aspect)*sin(slope)*sin(latitude_radians) ) +
    ( (-1*0.331)*sin(latitude_radians)*sin(slope) ) + 
    ( 0.375*sin(folded_aspect)*sin(slope) )
  )
  names(hli) <- "hli"
  
  # # this doesn't work when used in terra::lapp ... but does when not in lapp?
  # hli <- terra::ifel(hli > 1, 1, hli)
  # hli <- terra::ifel(hli < 0, 0, hli)
  
  # # this works when used in terra::lapp ... but doesn't when not in lapp?
  # hli <- ifelse(hli > 1, 1, hli)
  # hli <- ifelse(hli < 0, 0, hli)
  
  return(hli)
  
}
# # testtest
# get_hli_fn(
#   latitude_radians = lat_rad_rast[1:100, 1:100, drop = FALSE]
#   , slope = terrain_rad_rast$slope[1:100, 1:100, drop = FALSE]
#   , aspect = terrain_rad_rast$aspect[1:100, 1:100, drop = FALSE]
# ) %>% 
# terra::plot()


# do it with the hli
# identical(terra::ext(terrain_rad_rast$slope), terra::ext(dem_1m_rast))
# terra::ext(terrain_rad_rast$slope)
# terra::ext(dem_1m_rast)
# ?terra::extend
# identical(terra::ext(terrain_rad_rast$slope), terra::ext(dem_1m_rast))
f5_temp <- file.path(dirname(dem_fnm), "hli_1m_rast.tif" )
if(!file.exists(f5_temp)){

  # combine all required inputs into stack
  # make sure they have names to match function arguments
  # make sure they are in the order of the function arguments
  rast_stack_temp <- c(lat_rad_rast, terrain_rad_rast$slope, terrain_rad_rast$aspect)
  names(rast_stack_temp) <- c("latitude_radians", "slope", "aspect")
  
  # execute directly to disk
  # ?terra::lapp
  hli_1m_rast <- terra::lapp(
    rast_stack_temp
    , fun = get_hli_fn
    # # , get_hli_as_index = T
    , usenames = T
    
    , cores = lasR::half_cores()
    , filename = f5_temp
    , overwrite = F
  )

  
    
  # ############################################
  # # this version does math in memory
  # hli_1m_rast <- get_hli_fn(
  #   latitude_radians = lat_rad_rast
  #   , slope = terrain_rad_rast$slope # terra::terrain(dem_1m_rast, v = "slope", unit = "radians")
  #   , aspect = terrain_rad_rast$aspect # terra::terrain(dem_1m_rast, v = "aspect", unit = "radians")
  #   , get_hli_as_index = T
  # ) 
  # # write
  # terra::writeRaster(
  #   hli_1m_rast
  #   , filename = f5_temp, overwrite = F
  # )
  # ############################################
  # # this version does math in memory
  
}else{
  hli_1m_rast <- terra::rast(f5_temp)
}
# terra::minmax(hli_1m_rast)
hli_1m_rast <- terra::ifel(hli_1m_rast > 1, 1, hli_1m_rast)
hli_1m_rast <- terra::ifel(hli_1m_rast < 0, 0, hli_1m_rast)
# terra::minmax(hli_1m_rast)
names(hli_1m_rast) <- "hli"

terra::plot(shade_rast, col = grDevices::gray(0:100/100), legend = F, axes = F, mar=c(1,1,1.6,4))
terra::plot(
  hli_1m_rast
  , col = viridis::cividis(100)
  , range = c(0, 1), fill_range = T
  , add = T
  , alpha = 0.5
  , plg = list(
     title = "HLI"
     , x = "right"
     , shrink = 0.6
     , title.cex = 0.8
     , cex = 0.8
   )
)
terra::plot(terra::vect(cheesman_sf), col = NA, border = "black", add = T)
```

looks pretty hot. what is the distribution of HLI over this landscape in view?

```{r, warning=F}
terra::hist(hli_1m_rast, main = "HLI")
```

let's make the data more coarse which might help us better visualize high versus low HLI areas

```{r, results=F, fig.show='asis', fig.height=9, fig.width=5.5}
terra::plot(shade_rast, col = grDevices::gray(0:100/100), legend = F, axes = F, mar=c(1,1,1.6,4))
# ?terra::aggregate
terra::aggregate(hli_1m_rast, fact = 20, fun = "median", na.rm = T, cores = lasR::half_cores()) %>% 
terra::plot(
  col = viridis::cividis(100)
  , range = c(0, 1), fill_range = T
  , add = T
  , alpha = 0.5
  , plg = list(
     title = "HLI"
     , x = "right"
     , shrink = 0.6
     , title.cex = 0.8
     , cex = 0.8
   )
)
terra::plot(terra::vect(cheesman_sf), col = NA, border = "black", add = T)
```

what if we group by those HLI classes we noted above?

```{r, results=F, fig.show='asis', fig.height=9, fig.width=5.5}
# viridis::cividis(16) %>% scales::show_col()
pal_hli_grp <- viridis::cividis(16)[c(1,5,12,16)]
# pal_hli_grp %>% scales::show_col()

terra::plot(shade_rast, col = grDevices::gray(0:100/100), legend = F, axes = F, mar=c(1,1,2,6))
# ?terra::aggregate
hli_20m_rast_temp <- terra::aggregate(hli_1m_rast, fact = 20, fun = "median", na.rm = T, cores = lasR::half_cores())
hli_20m_rast_temp %>% 
terra::plot(
  col = pal_hli_grp
  , breaks = c(0,hli_grp_df$hli_ul)
  , add = T
  , alpha = 0.5
  , plg = list(
    legend = hli_grp_df$hli_grp
    , col = pal_hli_grp
     , title = "HLI Class"
     # , x = "top"
     # , shrink = 0.6
     , title.cex = 0.8
     , cex = 0.8
   )
)
terra::plot(terra::vect(cheesman_sf), col = NA, border = "black", add = T)
```

yeah, not many cool-mesic areas

we can check out the distribution of HLI aggregated to 20m cells by those groupings

```{r}
hli_20m_rast_temp %>% 
  terra::classify(rcl=c(0,hli_grp_df$hli_ul), include.lowest=F) %>% 
  terra::freq() %>% 
  dplyr::select(-layer) %>% 
  dplyr::rename(hli=1) %>% 
  dplyr::mutate(
    hli_ul = 
      stringr::str_sub(hli, start = stringr::str_locate(hli,"-")[,1]+1, end = stringr::str_length(hli)-1 ) %>%
      stringr::str_squish() %>%
      as.numeric()
  ) %>% 
  dplyr::left_join(hli_grp_df %>% dplyr::select(hli_ul,hli_grp)) %>% 
  dplyr::select(-hli_ul) %>% 
  dplyr::relocate(hli,hli_grp) %>% 
  dplyr::mutate(
    percent = scales::percent(count/sum(count), accuracy=0.1)
    , count = scales::comma(count, accuracy=1)
  ) %>% 
  dplyr::rename(
    `HLI Range`=hli
    , `HLI Class`=hli_grp
  ) %>% 
  kableExtra::kbl(caption = "distribution of HLI class in Cheesman Reservoir landscape") %>% 
  kableExtra::kable_styling()
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Surviving Tree Candidate Areas

We used satellite imagery to manually annotate point locations of surviving trees by planting year in planting locations where we felt the most confident in the year and area of the planting. For each planting year, we attempted to manually identify at least 10 surviving trees. Recall that our sampling strategy is presence-focused (i.e. non-random) because we do not know exactly where historical planting occurred, focusing on surviving clusters guarantees that crews are sampling from areas that were actually planted.

let's check those candidate sample plot locations centered on surviving tree points

```{r}
# I went about to getting at least 10 trees per planting year. 
# I tried to capture areas that encompass the entire planting areas for each planting year,
# but trees were admittedly a little sparse, 
# which made me make it a bit more dense in areas where there was good survival. 
known_trees_sf <-
  readxl::read_excel("../data/Trees_PlantingYears0406070913.xlsx") %>% 
  dplyr::rename_with(tolower) %>% 
  dplyr::filter(!is.na(plantingyear), !is.na(lat), !is.na(long)) %>% 
  sf::st_as_sf(coords = c("long","lat"), crs = sf::st_crs(4326), remove=F) %>%  
  sf::st_transform(terra::crs(dem_1m_rast)) %>% 
  dplyr::rename(planting_year=plantingyear) %>% 
  dplyr::group_by(planting_year) %>% 
  dplyr::mutate(
    year_tree_id = dplyr::row_number()
    , year_tree_fact = forcats::fct_cross(
      factor(planting_year)
      , factor(year_tree_id)
      ,sep = "_"
    )
  ) %>% 
  dplyr::ungroup()
# known_trees_sf %>% dplyr::glimpse()
```

where are our candidate plot centers (i.e. known tree locations)?

```{r, results=F, fig.show='asis', fig.height=9, fig.width=5.5}
# known_trees_sf %>% dplyr::glimpse()
# plot
terra::plot(shade_rast, col = grDevices::gray(0:100/100), legend = F, axes = F, mar=c(1,1,1.6,4))
terra::plot(
  dem_1m_rast
  , col = harrypotter::hp(n=100, option = "mischief")
  , add = T
  , alpha = 0.6
  , plg = list(
     title = "elevation (m)"
     , x = "right"
     , shrink = 0.6
     , title.cex = 0.8
     , cex = 0.8
   )
)
terra::plot(terra::vect(cheesman_sf), col = NA, border = "black", add = T)
terra::plot(terra::vect(known_trees_sf), col = "cyan", cex = 0.6, add = T)

# terra::plot(terra::vect(known_trees_sf), col = "cyan")
# terra::plot(dem_1m_rast, alpha = 0.3, add=T)
# terra::plot(terra::vect(cheesman_sf), col = NA, border = "black", add = T)

```

let's put those on an interactive map

```{r}
known_trees_sf %>% 
  dplyr::mutate(planting_year=factor(planting_year)) %>% 
  mapview::mapview(zcol = "planting_year", layer.name="known trees")
```

nice. let's buffer each potential plot location to create square areas of 0.5 and 1 acres

```{r}
make_square_plots_fn <- function(sf, plot_acres) {
  if (!inherits(sf, "sf")) {
    stop("sf must be an sf object.")
  }

  if(
    !is.numeric(plot_acres) ||
    length(plot_acres) != 1 ||
    is.na(plot_acres) ||
    plot_acres <= 0
  ){
    stop("plot_acres must be a single positive numeric value")
  }

  if(sf::st_is_longlat(sf)) {
    stop("sf must use a projected CRS with linear units as meters")
  }

  # Convert acres to square meters
  area_m2 <- plot_acres * 4046.8564224

  # Square side length, then half-side for the buffer radius
  radius_m <- sqrt(area_m2) / 2

  # Create square polygons centered on each point
  sf::st_buffer(
    sf
    , dist = radius_m
    , endCapStyle = "SQUARE"
  )
}
# make a few options
candidate_plots_sf <- dplyr::bind_rows(
  known_trees_sf %>% make_square_plots_fn(0.5) %>% dplyr::mutate(plot_area_ac = 0.5)
  , known_trees_sf %>% make_square_plots_fn(1.0) %>% dplyr::mutate(plot_area_ac = 1.0)
)

# dplyr::glimpse(candidate_plots_sf)
# candidate_plots_sf %>% 
#   dplyr::mutate(
#     area_m2 = sf::st_area(.) %>% as.numeric()
#     , area_ac = area_m2/4046.86
#   ) %>% 
#   sf::st_drop_geometry() %>% 
#   dplyr::mutate(dplyr::across(tidyselect::contains("area_"), ~round(.x,2) %>% scales::comma(accuracy = 0.01) )) %>% 
#   dplyr::count(plot_area_ac, area_ac, area_m2)
```

We'll use the following three topographic variables to capture an independent set of environmental conditions:

* Elevation: Controls for elevational shifts, such as temperature and precipitation
* Slope: Controls for structural physical dynamics, such as gravity-driven water drainage, soil erosion rates, and UAS sensor orientation/stretching during imagery capture.
* HLI: Controls for localized solar exposure and moisture deficits

```{r, results=F}
# stack
terrain_stack_1m_temp <- c(dem_1m_rast, terrain_rad_rast$slope, hli_1m_rast)
names(terrain_stack_1m_temp) <- c("elevation", "slope", "hli")
# terrain_stack_1m_temp

# exactextractr::exact_extract for zonal stats
candidate_plots_zonal <-
  exactextractr::exact_extract(
    x = terrain_stack_1m_temp
    , y = candidate_plots_sf
    , fun = c("mean", "stdev")
    , append_cols = c("year_tree_fact", "plot_area_ac")
  ) %>% 
  dplyr::rename_with(~stringr::str_replace_all(.x,"\\.","_")) %>% 
  # drop elevation stdev since it is collinear with mean slope
  dplyr::select(-stdev_elevation)
# add geoms
candidate_plots_zonal <- candidate_plots_sf %>% 
  dplyr::inner_join(
    candidate_plots_zonal
    , by = dplyr::join_by(year_tree_fact, plot_area_ac)
  )
# candidate_plots_zonal %>% dplyr::glimpse()
```

let's visualize those environmental values for the 1.0 acre candidate plots for a single planting year unit as an example

```{r,eval=T}
ex_temp <- 
  candidate_plots_zonal %>% 
  dplyr::filter(
    plot_area_ac==1.0
    , planting_year == known_trees_sf$planting_year[1]
  )
aoi_temp <- ex_temp %>% sf::st_buffer(22) %>% sf::st_union() %>% sf::st_bbox() %>% sf::st_as_sfc()
# ggplot2::ggplot() +
#   # cheesman
#   ggplot2::geom_sf(
#     data = cheesman_sf %>% sf::st_intersection(aoi_temp)
#     , fill = NA, color = "black"
#   ) 

# candidate_plots_zonal %>% 
#   dplyr::mutate(mean_slope = mean_slope*(180/pi)) %>% 
#   dplyr::select(year_tree_id, tidyselect::starts_with("mean_")) %>% 
#   tidyr::pivot_longer(cols = tidyselect::starts_with("mean_"), names_prefix = "mean_")

# rast
shade_rast_temp <- shade_rast %>% 
  terra::crop(
    aoi_temp %>% terra::vect()
    , mask = T
  ) %>% 
  terra::aggregate(2)
dem_1m_rast_temp <- dem_1m_rast %>% 
  terra::crop(
    aoi_temp %>% terra::vect()
    , mask = T
  ) %>% 
  terra::aggregate(2)
# plot  
plt_bg_temp <- 
  ggplot2::ggplot() +
  ggplot2::geom_tile(
    data = shade_rast_temp %>% 
      terra::as.data.frame(xy = T) %>% 
      dplyr::rename(f=3)
    , mapping = ggplot2::aes(x=x, y=y, fill = f)
  ) +
  ggplot2::scale_fill_distiller(palette = "Greys", direction = -1, na.value = "transparent", guide = "none") +
  #elevation
  ggnewscale::new_scale_fill() +
  ggplot2::geom_tile(
    data = dem_1m_rast_temp %>% 
      terra::as.data.frame(xy = T) %>% 
      dplyr::rename(f=3)
    , mapping = ggplot2::aes(x=x, y=y, fill = f)
    , alpha = 0.5
  ) +
  harrypotter::scale_fill_hp(
    "mischief"
    , na.value = "transparent"
    , name = "elevation (m)"
    , limits = terra::minmax(dem_1m_rast) %>% c() %>% `*`(0.95)
    , guide = "none"
  ) +
  ggplot2::scale_x_continuous(expand = c(0, 0)) +
  ggplot2::scale_y_continuous(expand = c(0, 0)) +
  ggplot2::labs(
    x = ""
    , y = ""
    , title = paste0(
      "Environmental conditions in example candidate plots of "
      , scales::comma(ex_temp$plot_area_ac[1],accuracy = 0.1, suffix = "ac")
      , " from the "
      , ex_temp$planting_year[1]
      , " planting year "
    )
  ) +
  ggplot2::theme_void() +
  ggplot2::theme(
    legend.position = "top" # c(0.5,1)
    , legend.direction = "horizontal"
    , legend.margin = ggplot2::margin(5,0,0,0)
    , legend.text = ggplot2::element_text(size = 6)
    , legend.title = ggplot2::element_text(size = 6)
    , legend.key = ggplot2::element_rect(fill = NA, color = NA)
    , plot.title = ggplot2::element_blank() # ggplot2::element_text(size = 10, face = "bold")
    , plot.subtitle = ggplot2::element_text(size = 8, hjust = 0.5, face = "bold")
  )
##### slope
plt_slope_temp <- 
  plt_bg_temp +
  # ggplot2::ggplot() +
  # # cheesman
  # ggplot2::geom_sf(
  #   data = cheesman_sf %>% sf::st_intersection(aoi_temp)
  #   , fill = NA, color = "black"
  # ) +
  # plots
  ggnewscale::new_scale_fill() +
  ggplot2::geom_sf(
    data = ex_temp %>%  dplyr::mutate(mean_slope = mean_slope*(180/pi))
    , mapping = ggplot2::aes(fill = mean_slope)
    , color = "gray66"
    , alpha = 0.7
  ) +
  ggplot2::scale_fill_viridis_c(
    option = "inferno"
    , limits = c(
      min(candidate_plots_zonal$mean_slope)
      , max(candidate_plots_zonal$mean_slope)
    ) %>% `*`((180/pi))
      # terrain_rad_rast$slope %>% 
      # terra::aggregate(60) %>% 
      # terra::minmax() %>% 
      # c() %>% 
      # `*`((180/pi))
    , name = "mean slope (deg)"
  ) +
  ggplot2::labs(subtitle = "Slope")

##### elevation
plt_elevation_temp <-
  plt_bg_temp +
  # ggplot2::ggplot() +
  # # cheesman
  # ggplot2::geom_sf(
  #   data = cheesman_sf %>% sf::st_intersection(aoi_temp)
  #   , fill = NA, color = "black"
  # ) +
  # plots
  ggnewscale::new_scale_fill() +
  ggplot2::geom_sf(
    data = ex_temp
    , mapping = ggplot2::aes(fill = mean_elevation)
    , color = "gray66"
    , alpha = 0.7
  ) +
  harrypotter::scale_fill_hp(
    option = "mischief"
    , limits = c(
      min(candidate_plots_zonal$mean_elevation)
      , max(candidate_plots_zonal$mean_elevation)
    )
      # terrain_rad_rast$elevation %>% 
      # terra::aggregate(60) %>% 
      # terra::minmax() %>% 
      # c() %>% 
      # `*`((180/pi))
    , name = "mean elevation (m)"
    , guide = ggplot2::guide_colorbar( theme = ggplot2::theme(legend.text = ggplot2::element_text(angle = 45, hjust = 1)) )
  ) +
  ggplot2::labs(subtitle = "Elevation")

##### hli
plt_hli_temp <-
  plt_bg_temp +
  # ggplot2::ggplot() +
  # # cheesman
  # ggplot2::geom_sf(
  #   data = cheesman_sf %>% sf::st_intersection(aoi_temp)
  #   , fill = NA, color = "black"
  # ) +
  # plots
  ggnewscale::new_scale_fill() +
  ggplot2::geom_sf(
    data = ex_temp
    , mapping = ggplot2::aes(fill = mean_hli)
    , color = "gray66"
    , alpha = 0.7
  ) +
  ggplot2::scale_fill_viridis_c(
    option = "cividis"
    , limits = c(0, 1)
    , name = "mean HLI"
    # , guide = ggplot2::guide_colorbar( theme = ggplot2::theme(legend.text = ggplot2::element_text(angle = 45, hjust = 1)) )
  ) +
  ggplot2::labs(subtitle = "HLI")

# patchwork
plt_temp <- 
  patchwork::wrap_plots(
    plt_elevation_temp, plt_slope_temp, plt_hli_temp
    , ncol = 1
  )
```

```{r,eval=T, fig.height = 11, fig.width = 6.6}
print(plt_temp)
ggplot2::ggsave("../data/ex_environmental_covariates.jpg", plot = plt_temp, height = 11, width = 6.6)
```

neat. even within a planting year area there is some variation in the environmental variables describing the candidate plot locations

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Identify Similar Candidate Plot Locations

To isolate how well the UAS workflow performs across different tree sizes and ages (planting years) using the field validation data, we need to establish matching topographic plot locations across the planting years. By selecting sample plot locations (non-random) that are topographically similar, we can better isolate the independent variable we are seeking to monitor — tree presence and structure as measured by the UAS — from sensor artifacts caused by topography (e.g., slope steepness, lighting conditions like shadows and glare).

To control for terrain-based UAS sensor changes and generate an unbiased validation data set of field-measured values, we'll use a statistical — *not a spatial* — Euclidean nearest-neighbor matching process to create plot sets by calculating the SSE across multidimensional covariates (i.e. elevation, slope, and HLI). By minimizing the SSE across the possible planting year plot combinations, this workflow groups the candidate plot polygons into matching/similar topographic cohorts.

```{r}
get_best_match_set_temp <- function(my_plot_area_ac) {
  # which
  which_df_temp <- 
    candidate_plots_zonal %>% 
    dplyr::filter(plot_area_ac==my_plot_area_ac)
  # select covariates and scale
  plots_temp <- 
    which_df_temp %>% 
    sf::st_drop_geometry() %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(
      dplyr::across(
        .cols = c(tidyselect::starts_with("mean_"),tidyselect::starts_with("stdev_"))
        , .fns  = ~ base::scale(.x) %>% as.numeric()
      )
    )
  # plots_temp %>% dplyr::glimpse()
  # plots_temp %>% dplyr::select(c(tidyselect::starts_with("mean_"),tidyselect::starts_with("stdev_"))) %>% summary()
  
  # compare all to one year
  first_yr_temp <- min(plots_temp$planting_year)
  
  first_yr_plots_temp <- 
    plots_temp %>% 
    dplyr::filter(planting_year == first_yr_temp) %>% 
    dplyr::rename_with(
      .cols = c(year_tree_fact,year_tree_id,planting_year)
      , ~paste0("anchor_", .x, recycle0 = T)
    ) %>% 
    dplyr::select(
      tidyselect::starts_with("anchor_")
      , tidyselect::starts_with("mean_"),tidyselect::starts_with("stdev_")
    ) %>% 
    sf::st_drop_geometry()
  # first_yr_plots_temp %>% dplyr::glimpse()
  
  comp_plots_temp <- plots_temp %>% 
    dplyr::filter(planting_year != first_yr_temp) %>% 
    dplyr::select(
      year_tree_fact,year_tree_id,planting_year
      , tidyselect::starts_with("mean_"),tidyselect::starts_with("stdev_")
    ) %>% 
    dplyr::rename_with(~paste0("comp_", .x, recycle0 = T)) %>% 
    sf::st_drop_geometry()
  # comp_plots_temp %>% dplyr::glimpse()
  
  # calc SSE across all pairwise combinations and pick the lowest sse
  matched_pairs_temp <-
    first_yr_plots_temp %>% 
    tidyr::crossing(comp_plots_temp) %>% 
    # dplyr::glimpse()
    # Calculate squared differences for each covariate pair
    dplyr::mutate(
      se_elev = (mean_elevation - comp_mean_elevation)^2
      , se_mslp = (mean_slope - comp_mean_slope)^2
      , se_sslp = (stdev_slope - comp_stdev_slope)^2
      , se_mhli = (mean_hli - comp_mean_hli)^2
      , se_shli = (stdev_hli - comp_stdev_hli)^2
      # sum of squared errors (sse)
      , sse_total = se_elev + se_mslp + se_sslp + se_mhli + se_shli
    ) %>% 
    # get a ranking of the most similar for each anchor plot,year
    dplyr::group_by(anchor_year_tree_fact,anchor_year_tree_id,anchor_planting_year,comp_planting_year) %>% 
    dplyr::slice_min(order_by = sse_total, n = 1, with_ties = FALSE) %>% 
    dplyr::ungroup() %>% 
    # sum the sse by anchor tree, then pick the min sse overall
    dplyr::group_by(anchor_year_tree_fact,anchor_year_tree_id,anchor_planting_year) %>% 
    dplyr::mutate(set_sse_total = sum(sse_total)) %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(rank_sse = dplyr::dense_rank(set_sse_total)) %>% 
    dplyr::arrange(rank_sse,anchor_year_tree_fact,comp_planting_year)
  # matched_pairs_temp %>% dplyr::glimpse()
  
  # pick the min sse overall
  best_year_tree_fact_temp <-
    matched_pairs_temp %>% 
    dplyr::filter(
      set_sse_total == min(set_sse_total)
    ) %>% 
    dplyr::select(tidyselect::ends_with("year_tree_fact")) %>% 
    tidyr::pivot_longer(dplyr::everything()) %>% 
    dplyr::pull(value) %>% 
    unique()
  
  # return
  return(list(
    candidate_plots_df = 
      which_df_temp %>% 
      dplyr::mutate(
        is_best_plot_set = ifelse(year_tree_fact %in% best_year_tree_fact_temp, T, F)
      )
    , matched_pairs_df = matched_pairs_temp
  ))
}
#map over plot sizes
best_match_set_list <- 
  candidate_plots_sf$plot_area_ac %>% 
  unique() %>% 
  purrr::set_names() %>% 
  purrr::map(get_best_match_set_temp)

# best_match_set_list[[1]]$matched_pairs_df %>% dplyr::glimpse()
# best_match_set_list[[1]]$candidate_plots_df %>% dplyr::glimpse()
# best_match_set_list %>% names()

matched_pairs_df <- best_match_set_list %>% purrr::map(~purrr::pluck(.x,"matched_pairs_df")) %>% purrr::list_rbind(names_to = "plot_area_ac")
# matched_pairs_df %>% dplyr::glimpse()
candidate_plots_zonal <-
  best_match_set_list %>% 
  purrr::map(~purrr::pluck(.x,"candidate_plots_df")) %>% 
  purrr::list_rbind() 
# candidate_plots_zonal  %>% 
#   sf::st_drop_geometry() %>%  
#   dplyr::count(plot_area_ac, is_best_plot_set)
# candidate_plots_zonal %>% dplyr::glimpse()
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

where are the selected plots in the topographic covariate space? Note, we have options for both 0.5 and 1.0 acre plot sizes depending on which is most feasible given time, cost, and workforce constraints

as a reminder of what we are looking at:

1) we manually annotated live tree point locations in each planting year location we want to sample and buffered these points to create candidate field plots (1- and 0.5-acre)
2) we used a 1-meter DEM to calculate five (non-collinear) terrain covariates for every candidate plot (1- and 0.5-acre): 
  - Mean Elevation 
  - Mean Slope 
  - Slope Standard Deviation 
  - Mean HLI (Heat Load Index)
  - HLI Standard Deviation
3) we normalized these covariates to ensure equal weighting (otherwise elevation would overwhelm the others simply based on the units)
4) using the earliest planting year as an anchor, we used a statistical nearest neighbor Euclidean matching process to search across all later planting years to select topographically similar plots that minimized the SSE across all five terrain covariates simultaneously

because the selected plot locations have similar topographic profiles, we have controlled for conditions, as best we can, which might influence the UAS sensor data collection. As a result, when we calculate our UAS validation metrics (F-score, RMSE, MAPE) based on the field data, the influence of slope steepness, terrain roughness, or shifting lighting conditions (shadows and glare) will be minimized.

```{r, fig.height=10, fig.width=9}
plt_matrix_temp <- function(my_plot_area_ac) {
  GGally::ggpairs(
    data = 
      candidate_plots_zonal %>% 
      dplyr::filter(plot_area_ac==my_plot_area_ac) %>% 
      sf::st_drop_geometry() %>% 
      dplyr::mutate(
        planting_year = factor(planting_year)
        , mean_slope = mean_slope*(180/pi)
      ) %>% 
      dplyr::arrange(is_best_plot_set)
    , columns = 
      # candidate_plots_zonal %>%
      # sf::st_drop_geometry() %>%
      # dplyr::select(tidyselect::starts_with("mean_"),tidyselect::starts_with("stdev_")) %>%
      # names()
      c("mean_elevation","mean_slope","mean_hli","stdev_slope","stdev_hli")
    
    , mapping = ggplot2::aes(
      color = is_best_plot_set 
      , fill  = is_best_plot_set 
      , shape = planting_year
    )
    , lower = list(continuous = GGally::wrap("points", size = 1.7, alpha = 0.9))
    , upper = "blank"
    , axisLabels = "show"
    , switch     = "both"
    , showStrips = T
    , columnLabels =
      c(
        "Elevation mean (m)" # "mean_elevation"
        ,"Slope mean (deg)" # "mean_slope"
        ,"HLI mean (0-1)" # "mean_hli"
        ,"Slope SD (deg)" # "stdev_slope"
        ,"HLI SD (0-1)" # "stdev_hli"
      )
    , legend = c(2, 1) 
  ) +
    ggplot2::scale_color_manual(values = c("TRUE"="navy","FALSE"="gray"), name = "selected\nplot") +
    ggplot2::scale_fill_manual(values = c("TRUE"="navy","FALSE"="gray"), guide = "none") +
    ggplot2::scale_shape_manual(values = c(21:25), name = "planting\nyear") +
    ggplot2::labs(
      title = paste("Topographic Covariate Matrix for", my_plot_area_ac, "acre Plot Size")
    ) +
    ggplot2::theme_light() +
    ggplot2::theme(
      strip.text = ggplot2::element_text(size = 8, face = "bold", color = "black")
      # , strip.background = ggplot2::element_blank()
      , strip.placement = "outside"
      
      , legend.position = "top"
      , legend.direction = "horizontal"
      , legend.background = ggplot2::element_rect(fill = NA, color = "gray")
      , legend.key = ggplot2::element_blank()
      , legend.text = ggplot2::element_text(size = 8)
    ) +
    ggplot2::guides(
      color = ggplot2::guide_legend(
        override.aes = list(shape = 15, size = 6, linetype = 0)
      )
      , fill = "none"
    )  
}
# map
candidate_plots_zonal$plot_area_ac %>% 
  unique() %>% 
  sort(decreasing = T) %>% 
  purrr::map(plt_matrix_temp)

```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

finally, let's list the top 5 plot location sets (1 set = 1 plot per planting year) based on topographic similarity in case we need to shift plot locations

```{r}
matched_pairs_df %>% 
  dplyr::filter(rank_sse<=5) %>% 
  dplyr::select(plot_area_ac, anchor_year_tree_fact, comp_year_tree_fact, rank_sse, set_sse_total) %>% 
  tidyr::pivot_longer(tidyselect::ends_with("year_tree_fact"), values_to = "year_tree_fact") %>% 
  dplyr::distinct(plot_area_ac, year_tree_fact, rank_sse, set_sse_total) %>% 
  # add on lat/long e.g. 40.50321981785759, -104.99788267106209
  dplyr::inner_join(
    known_trees_sf %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(year_tree_fact, lat, long, planting_year)
    , by = "year_tree_fact"
    , relationship = "many-to-one" # bc multiple plot sizes
  ) %>% 
  # dplyr::glimpse()
  # make lat/long coordinates and pivot wider
  dplyr::mutate(
    coords = paste0(
      "("
      , scales::number(lat,accuracy = 0.000001)
      , ", "
      , scales::number(long,accuracy = 0.000001)
      , ")"
    )
    , set_sse_total = scales::number(set_sse_total,accuracy = 0.01)
  ) %>% 
  dplyr::select(-c(lat,long,year_tree_fact)) %>% 
  tidyr::pivot_wider(names_from = planting_year, values_from = coords) %>% 
  dplyr::arrange(plot_area_ac,rank_sse,set_sse_total) %>% 
  dplyr::rename(
    `Plot Area (ac)` = plot_area_ac
    , `Set Rank` = rank_sse
    , `SSE (normalized)` = set_sse_total
  ) %>% 
  kableExtra::kbl(
    caption = "top 5 plot location sets (1 set = 1 plot per planting year) based on topographic similarity"
  ) %>% 
  kableExtra::kable_styling(
    bootstrap_options = c("striped")
    , latex_options = c("striped")
  ) %>% 
  kableExtra::collapse_rows(1)
```

finally, let's look at those in space

*1.0 acre plots*

```{r}
# option to put satellite imagery as base layer of mapview maps
mapview::mapviewOptions(
  homebutton = FALSE
  , basemaps = c("Esri.WorldImagery","OpenStreetMap")
)
df_temp <- 
  candidate_plots_zonal %>% 
  dplyr::filter(
    plot_area_ac == 1
    , is_best_plot_set
  ) %>% 
  dplyr::mutate(
    planting_year=factor(planting_year)
    , coords = paste0(
      "("
      , scales::number(lat,accuracy = 0.000001)
      , ", "
      , scales::number(long,accuracy = 0.000001)
      , ")"
    )
  ) %>% 
  dplyr::select(planting_year,coords,lat,long,plot_area_ac,tidyselect::starts_with("mean_"),tidyselect::starts_with("stdev_"))
  
df_temp %>% 
  mapview::mapview(zcol = "planting_year", layer.name=paste0("selected plot locations: ",unique(df_temp$plot_area_ac)," ac"))
```

```{r, include=F, eval=T}
# df_temp %>% dplyr::glimpse()
# df_temp %>% sf::st_crs()
sf::st_write(df_temp, dsn = paste0("../data/selected_field_plots_",unique(df_temp$plot_area_ac),"ac.gpkg"), quiet = T, append = F)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

*0.5 acre plots*

```{r}
df_temp <- 
  candidate_plots_zonal %>% 
  dplyr::filter(
    plot_area_ac == 0.5
    , is_best_plot_set
  ) %>% 
  dplyr::mutate(
    planting_year=factor(planting_year)
    , coords = paste0(
      "("
      , scales::number(lat,accuracy = 0.000001)
      , ", "
      , scales::number(long,accuracy = 0.000001)
      , ")"
    )
  ) %>% 
  dplyr::select(planting_year,coords,lat,long,plot_area_ac,tidyselect::starts_with("mean_"),tidyselect::starts_with("stdev_"))
  
df_temp %>% 
  mapview::mapview(zcol = "planting_year", layer.name=paste0("selected plot locations: ",unique(df_temp$plot_area_ac)," ac"))
```

```{r, include=F, eval=T}
# df_temp %>% dplyr::glimpse()
# df_temp %>% sf::st_crs()
sf::st_write(df_temp, dsn = paste0("../data/selected_field_plots_",unique(df_temp$plot_area_ac),"ac.gpkg"), quiet = T, append = F)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```
